# Assignment 6: Analyzing longitudinal data

```{r}
library(readr)
library(dplyr)
library(GGally)
library(gridExtra)
```

## RATS data

>To begin to see how linear mixed effects models are applied in practice, we
shall use some data from a nutrition study conducted in three groups of rats
(Crowder and Hand, 1990). The three groups were put on different diets, and
each animal’s body weight (grams) was recorded repeatedly (approximately
weekly, except in week seven when two recordings were taken) over a 9-week
period. The question of most interest is whether the growth profiles of the
three groups differ. The data are shown in Table 9.1.

The RATS data is from a nutrition stufy by Crowdeer & Hand (1990) on three groups of rats, each group given a different diet.
The research question is whether the growth profiles of the rats differ.


## 1. Chapter 8 approach with rats dataset

>Implement the analyses of Chapter 8 of MABS, using the R codes of Exercise Set 6: Meet and Repeat: PART I but using the RATS data (from Chapter 9 and Meet and Repeat: PART II).
(0-7 points: 0-4 points for graphs or analysis results + 0-3 points for their interpretations)

In this problem, we're using the "summary measure approach from chapter 8".

```{r}
rats <- read_csv("data/rats.csv") %>% mutate(ID = factor(ID)) %>% mutate(Group = factor(Group))
```

### 1.1 Dataset {.tabset .tabset-fade }

Let's inspect:

 - the distribution of initial bodyweights
 - the distribution of final bodyweights
 - the distribution of bodyweights in the three groups
 - the mean growth profile
 - the growth profile of the three groups
 
Below, there are three tabs in the notebook, you can switch between them to look at different aspects of the data.

#### Distribution shift
 
We start by looking at the distribution of initial and final bodyweights to see if we can detect a trend. 

```{r fig.align="center", fig.cap = "fig. 1.1, Baseline and final rat weight distributions", fig.width = 16, fig.height = 16, results = 'hide'}
initial_time <- min(rats$Time)
final_time <- max(rats$Time)

initial_rats <- rats %>% filter(Time == initial_time)
final_rats <- rats %>% filter(Time == final_time)
min_range <- min(rats$Bodyweight)
max_range <- max(rats$Bodyweight)

ggplot() +    geom_density(data= initial_rats, bw = "SJ", aes(x=Bodyweight, fill = "Baseline bodyweight", alpha=0.4)) +
              geom_density(data= final_rats, bw = "SJ", aes(x=Bodyweight,  fill = "Final bodyweight", alpha=0.4)) +
              scale_fill_manual(values=c("Baseline bodyweight" = "#cccccc","Final bodyweight" = "#ff9955")) +
              labs(fill= "Weight distribution") +
              scale_x_continuous(limits = c(min_range, max_range)) +
              scale_alpha(guide = "none")

```
There is a clear shift towards a heavier distribution of bodyweights over the course of the experiment --- the rats are growing.
It looks like there's three modes in both distribution, but in the final bodyweight distribution, the middle mode has gotten close to the heaviest mode.

#### Groups

Let's look at the initial bodyweights of the three groups.

```{r fig.align="center", fig.cap = "fig. 1.2, Baseline weight distributions per group", fig.width = 16, fig.height = 16, results = 'hide'}
initial_rats %>%
  ggplot(aes(x=Bodyweight, group=Group, fill=Group, alpha = 0.5)) +
  geom_density(bw = "SJ")+
  scale_alpha(guide = "none")
```

Hmm, the groups seem to have different bodyweight distributions, the modes do not overlap, and I have a feeling that the difference between groups is much greater than the difference between the growth from initial to final measurements.
This is even clearer in the next tab to the right.

#### Distribution shift per group

```{r fig.align="center", fig.cap = "fig. 1.3, Baseline and final weight distributions per group", fig.width = 16, fig.height = 16, results = 'hide'}

ggplot() +    geom_density(data= initial_rats, bw = "SJ", aes(x=Bodyweight, group=Group, fill = paste("Group",Group,"baseline"), alpha=0.4)) +
              geom_density(data= final_rats, bw = "SJ", aes(x=Bodyweight, group=Group, fill = paste("Group",Group,"final"), alpha=0.4)) +
              scale_fill_manual(values=c("Group 1 baseline" = "#ff9999",
                                         "Group 2 baseline" = "#99ff99",
                                         "Group 3 baseline" = "#9999ff",
                                         "Group 1 final" = "#dd2222",
                                         "Group 2 final" = "#22dd22",
                                         "Group 3 final" = "#2222dd")) +
              labs(fill= "Weight distribution") +
              scale_x_continuous(limits = c(min_range, max_range)) +
              scale_alpha(guide = "none")
```

The shift of the distributions are visible, but they are still smaller than the initial differences between the weight distributions.
In my mind this would make the groups a little difficult to compare, since we can't be sure that the effect of the baseline bodyweight doesn't behave differently for each group.
The green high peak is a single outlier, as we will see when looking at the line plots.

#### Growth trend

```{r fig.align="center", fig.cap = "fig. 1.4, Mean growth and plot of each rats bodyweight increase", fig.width = 16, fig.height = 16, results = 'hide'}
rat_trend <-  rats %>% group_by(Time) %>%
  summarise(mean = mean(Bodyweight), se = sd(Bodyweight)/sqrt(n())) %>%
  ungroup()

rat_trend %>% ggplot(aes(x=Time, y = mean)) +
          geom_line(aes(color="mean")) +
          geom_errorbar(aes(ymin=mean-se, ymax=mean+se, color = "mean"), width=0.3) +
          labs(color= "Group") +
          ylab("Bodyweight") +
          geom_line(data=rats, aes(x=Time, y=Bodyweight, color = Group, group=ID))
```

We see an issue with taking the mean bodyweight of all rats: none of the rats are at the mean, they straddle the mean on both sides.
We also see more clearly in this plot that one of the groups (group 1) has double the number of rats that the others do (8 vs. 4).
The outlier in the green group is also much more apparent here, we might have to remove it.

### 1.2 Standardize

Let's standardize the dataset and plot that last plot (`Growth Trend`) again.

```{r fig.align="center", fig.cap = "fig. 1.5, Mean time-standardized weights and plot of each rats std weight", fig.width = 16, fig.height = 16, results = 'hide'}
rats_std <- rats %>%
  group_by(Time) %>%
  mutate(std_weight = scale(Bodyweight)) %>%
  ungroup()

std_rat_trend <-  rats_std %>% group_by(Time) %>%
  summarise(mean = mean(std_weight), se = sd(std_weight)/sqrt(n())) %>%
  ungroup()

std_rat_trend %>% ggplot(aes(x=Time, y = mean)) +
          geom_line(aes(color="mean")) +
          geom_errorbar(aes(ymin=mean-se, ymax=mean+se, color = "mean"), width=0.3) +
          labs(color= "Group") +
          ylab("standardized weight") +
          geom_line(data=rats_std, aes(x=Time, y=std_weight, color = Group, group=ID))

```


### 1.3 mean plots

There are so few samples per group that the figure 1.5 contains more information than plotting the mean and standard error, but let's plot that as well for completeness.

```{r fig.align="center", fig.cap = "fig. 1.6, Mean growth per group", fig.width = 16, fig.height = 16, results = 'hide'}

std_rat_group_trends <-  rats_std %>% group_by(Time, Group) %>%
  summarise(mean = mean(std_weight), se = sd(std_weight)/sqrt(n())) %>%
  ungroup()

std_rat_group_trends %>% ggplot(aes(x = Time, y = mean, color = Group)) +
  geom_line() +
  geom_point(size=3) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=0.3) +
  scale_y_continuous(name = "mean(Bodyweight) +/- se(Bodyweight)")
```

### 1.4 Outlier in group 2?

### 1.5 Standardize by baseline?

### 1.6 fit and anova?

## 2. BPRS and chapter 9 analysis

>Implement the analyses of Chapter 9 of MABS, using the R codes of Exercise Set 6: Meet and Repeat: PART II, but using the BPRS data (from Chapter 8 and Meet and Repeat: PART I).
(0-8 points: 0-4 points for graphs or analysis results + 0-4 points for their interpretations)

